# Introduction to R and R Studio

## Introduction to R 

Much of your time in R will be spent in the R interactive console. This is where you will run all of your code, and can be a useful environment to try out ideas before adding them to an R script file. This console in RStudio is the same as the one you would get if you typed in R in your command-line environment.

The first thing you will see in the R interactive session is a bunch of information, followed by a “>” and a blinking cursor. In many ways this is similar to the shell environment you learned about during the shell lessons: it operates on the same idea of a “Read, evaluate, print loop”: you type in commands, R tries to execute them, and then returns a result.

## Using R as a calculator

The simplest thing you could do with R is do arithmetic:

```{r}
1+100
```

R will print out the answer with a preceding ```[1]```. Don't worry about this for now, we'll explain that later. For now, think of it as indicating output.

As with bash, if you type an incomplete command, R will wait for you to complete it:

```{r, eval=FALSE}
1+
```

Any time you hit return and the R session show a ```+``` instead of a ```>``` it means it's waiting for you to complete the command. If you want to cancel a command, you can simply hit ```ESC``` and RStudio will give you back the ">" prompt. 

When using R as a calculator, the order of operations is the same as you would have learned back in school. 

From highest to lowest precedence:

Parentheses: (, )
Exponents: ^ or **
Divide: /
Multiply: *
Add: +
Subtract: -

```{r}
3+5*2
```

```{r}
(3+5)*2
```

```{r}
(3 + (5 * (2 ^ 2))) # hard to read
3 + 5 * 2 ^ 2       # clear, if you remember the rules
3 + 5 * (2 ^ 2)     # if you forget some rules, this might help
```

The text after each line of code is called a “comment”. Anything that follows after the hash (or octothorpe) symbol # is ignored by R when it executes code.

Really small or large numbers get a scientific notation:

```{r}
2/10000
```

This is shorthand for "multiplied by ```10^XX```. So ```2e-4``` is shorthand for ```2*10^(-4)```.

You can write numbers in scientific notation too:

```{r}
5e3
```

Don’t worry about trying to remember every function in R. You can look them up on Google, or if you can remember the start of the function’s name, use the tab completion in RStudio.

This is one advantage that RStudio has over R on its own, it has auto-completion abilities that allow you to more easily look up functions, their arguments, and the values that they take.

Typing a ? before the name of a command will open the help page for that command. As well as providing a detailed description of the command and how it works, scrolling to the bottom of the help page will usually show a collection of code examples which illustrate command usage. We’ll go through an example later.


## Comparing things

We can also do comparison in R:

```{r}
1==1 # equality; note two equals signs, read as 'is equal to'
```

```{r}
1 != 2 # inequality (read as 'is not equal to')
```

```{r}
1<2 # less than
```

```{r}
1<=1 # less than or equal to
```

```{r}
1>0 # greater than
```

```{r}
1 >= -9 # greater than or equal to
```

## Variables and assignment

We can store values in variables using the assignment operator ```<-``` like this:

```{r}
x<-1/40
```

This operation assigned the value 1/40 to the variable x. Notice that assignment does not print a value. Instead, we stored it for later in the variable x, which is now associated with 1/40. 

You can confirm this by printing x, which returns the value 0.025:

```{r}
x
```

More More precisely, the stored value is a decimal approximation of this fraction called a floating point number.

Look for the Environment tab in one of the panes of RStudio, and you will see that x and its value have appeared. Our variable x can be used in place of a number in any calculation that expects a number:

```{r}
log(x)
```

Note also that variables can be reassigned:

```{r}
x<-100
```

x used to contain the value 0.025 and and now it has the value 100.

Assignment values can contain the variable being assigned to:

```{r}
x <- x + 1 #notice how RStudio updates its description of x on the top right tab
y <- x * 2
```

The right hand side of the assignment can be any valid R expression. The right hand side is fully evaluated before the assignment occurs.

Variable names can contain letters, numbers, underscores and periods. They cannot start with a number nor contain spaces at all. Different people use different conventions for long variable names, these include

periods.between.words
underscores_between_words
camelCaseToSeparateWords
What you use is up to you, but be consistent.

It is also possible to use the = operator for assignment:

```{r}
x=1/40
```

But this is much less common among R users. The most important thing is to be consistent with the operator you use. There are occasionally places where it is less confusing to use <- than =, and it is the most common symbol used in the community. So the recommendation is to use <-.

## Installing packages

We can use R as a calculator to do mathematical operations (e.g., addition, subtraction, multiplication, division), as we did above. However, we can also use R to carry out more complicated analyses, make visualizations, and much more. In later episodes, we’ll use R to do some data wrangling, plotting, and saving of reformatted data.

R coders around the world have developed collections of R code to accomplish themed tasks (e.g., data wrangling). These collections of R code are known as R packages. It is also important to note that R packages refer to code that is not automatically downloaded when we install R on our computer. Therefore, we’ll have to install each R package that we want to use (more on this below).

We will practice using the dplyr package to wrangle our datasets in episode 6 and will also practice using the ggplot2 package to plot our data in episode 7. To give an example, the dplyr package includes code for a function called filter(). A function is something that takes input(s) does some internal operations and produces output(s). For the filter() function, the inputs are a dataset and a logical statement (i.e., when data value is greater than or equal to 100) and the output is data within the dataset that has a value greater than or equal to 100.

There are two main ways to install packages in R:

1. If you are using RStudio, we can go to Tools > Install Packages... and then search for the name of the R package we need and click Install.

2. We can use the ```install.packages( )``` function. We can do this to install the dplyr R package.

```{r, eval=FALSE}
install.packages("dplyr")
```


It’s important to note that we only need to install the R package on our computer once

Now that we’ve installed the R package, we’re ready to use it! To use the R package, we need to “load” it into our R session. We can think of “loading” an R packages as telling R that we’re ready to use the package we just installed. It’s important to note that while we only have to install the package once, we’ll have to load the package each time we open R (or RStudio).

To load an R package, we use the library( ) function. We can load the dplyr package like this:

```{r, eval=FALSE}
library(dplyr)
```


