# Exploring Data Frames

At this point, you’ve seen it all: in the last lesson, we toured all the basic data types and data structures in R. Everything you do will be a manipulation of those tools. But most of the time, the star of the show is the data frame—the table that we created by loading information from a csv file. In this lesson, we’ll learn a few more things about working with data frames.

## Realistic Example

We already learned that the columns of a data frame are vectors, so that our data are consistent in type throughout the columns. So far, you have seen the basics of manipulating data frames with our nordic data; now, let's use those skills to analyze a more complext dataset.

Let's go ahead and read in the gapminder dataset and assign it to an object named "gapminder"

```{r, echo=-1}
setwd("/Users/adra7980/Documents/carpentries_demo/data")
gapminder<-read.csv("gapminder_data.csv")
```

Let's investigate the gapminder data frame a bit; the first thing we should always do is get a sense of what the data looks like with the ```str``` function: 

```{r}
str(gapminder)
```

We can also examine individual columns of the data frame with our ```class``` function:

```{r}
class(gapminder$year)
```

```{r}
class(gapminder$country)
```

We can also apply ```str``` to individual columns

```{r}
str(gapminder$country)
```

We can also interrogate the data frame for information about its dimensions. Recall from earlier that when we passed gapminder to the ```str``` function, the output informed us that there were 1704 observations and 6 variables in gapminder. Keeping that in mind what do you think the following will produce, and why?

```{r}
length(gapminder)
```

A fair guess would have been to say that the length of a data frame would be the number of rows (1704 in this case), but that's not the case; the length actually returns the number of columns. 

Let's confirm the class of ```gapminder```

```{r}
class(gapminder)
```

To get the number of rows in the dataset, we can use the ```nrow``` function:

```{r}
nrow(gapminder)
```

To get the number of columns, we can use the ```ncol``` function:

```{r}
ncol(gapminder)
```

If we want to get both at once, and get a fuller sense of a dataset's dimensions at one go, we can use the ```dim``` function:

```{r}
dim(gapminder)
```

We'll also likely want to know the titles of all the columns. We can use the ```colnames``` function to get those names:

```{r}
colnames(gapminder)
```

At this point, it's it’s important to ask ourselves if the structure R is reporting matches our intuition or expectations; do the basic data types reported for each column make sense? If not, we need to sort any problems out now before they turn into bad surprises down the road, using what we’ve learned about how R interprets data, and the importance of strict consistency in how we record our data.

Once we’re happy that the data types and structures seem reasonable, it’s time to start digging into our data proper. Check out the first few lines:

```{r}
head(gapminder)
```

## Adding columns and rows in data frames

Let's talk about adding columns and rows in data frames. Let's assume that we'd like to create a new column to hold information on whether the life expectancy is below or above the world's average life expectancy of 70.5:

```{r}
below_average<-gapminder$lifeExp<70.5
```

Let's refresh our memories about what gapminder looks like:

```{r}
head(gapminder)
```

And now, let's add ```below_average``` as a column using the ```cbind``` function:

```{r}
cbind(gapminder, below_average)
```

If we don't want to print the entire data frame every time, we can put our ```cbind``` command within a call to ```head``` to return only the first six lines of the output:

```{r}
head(cbind(gapminder, below_average))
```

Now let's see what happens if we try to add a vector of ```below_average``` with a different number of entries than the number of rows in the ```gapminder``` data frame:

```{r, eval=FALSE}
below_average<-c(TRUE, TRUE, TRUE, TRUE, TRUE)
head(cbind(gapminder, below_average))
```

Running this leads to an error that says "arguments imply differing number of rows." This happens because R wants to see one element in our new column for every row in the table. 

Let's remind ourselves of the number of rows in gapminder:

```{r}
nrow(gapminder)
```

And the length of the ```below_average``` vector:

```{r}
length(below_average)
```

For the binding to work, we need to have the number of rows in ```gapminder``` equal to the length of ```below_average``` or we need the number of rows in gapminder to be a multiple of the length of ```below_average```. Because ```gapminder``` does not have five rows or a multiple of five rows, the column binding failed. 

Let's try again, this time with a ```below_average``` vector length of three:

```{r}
below_average<-c(TRUE, TRUE, FALSE)
head(cbind(gapminder, below_average))
```

This time, it worked, since 1704 is divisible by three. More specifically, what's happening here is that the sequence ```TRUE, TRUE, FALSE``` is repeated over all the gapminder rows. Let's go ahead and overwrite the content of gapminder with this new data frame:

```{r}
below_average<-as.logical(gapminder$lifeExp<70.5)
gapminder<-cbind(gapminder, below_average)
```

Now how about adding rows? Recall that the rows of a data frame are lists. To add a new row, we can make a list and then use the ```rbind``` function to append the row to the existing data frame. Let's see an example:

```{r}
new_row<-list('Norway', 2016, 500000, 'Nordic', 80.3, 49400.0, FALSE)
gapminder_norway<-rbind(gapminder, new_row)
tail(gapminder_norway)
```

That worked, because ```continent``` is being treated as a character vector. But what if it was considered a factor?

```{r}
gapminder$continent<-as.factor(gapminder$continent)
new_row<-list('Norway', 2016, 500000, 'Nordic', 80.3, 49400.0, FALSE)
gapminder_norway<-rbind(gapminder, new_row)
tail(gapminder_norway)
```

You'll notice that this time, the binding still works, but R gives us a warning. To see what's going on, we need to learn a bit more about factors. 

## Factors

The thing to note here is that in factor, each different value represents what is called a level. In our case, the factor "continent" has 5 levels: "Africa", "Americas", "Asia", "Europe", and "Oceania." R will only accept values that match one of the levels. If you add a new row that has a different value for continent (such as Nordic in our case), it will become NA. 

More specifically, the warning was telling us that we unsuccessfully added “Nordic” to our continent factor, but 2016 (a numeric), 5000000 (a numeric), 80.3 (a numeric), 49400.0 (a numeric) and FALSE (a logical) were successfully added to country, year, pop, lifeExp, gdpPercap and below_average respectively, since those variables are not factors. ‘Norway’ was also successfully added since it corresponds to an existing level. 

To successfully add a gapminder row with a “Nordic” continent, we can add “Nordic” as a level of the factor. 

```{r}
# Add Nordic as a level of the factor
levels(gapminder$continent)<-c(levels(gapminder$continent), "Nordic")
# Confirm it's been added
levels(gapminder$continent)
# Add the row to gapminder
gapminder_norway<-rbind(gapminder,
                    list('Norway', 2016, 5000000, 'Nordic', 80.3,49400.0, FALSE))
```

Now, let's confirm that the row has been added and that the factor column doesn't cause problems:

```{r}
tail(gapminder_norway)
```

To successfully add the row, our options were to either work with the character vector, which makes adding rows less cumbersome, but doesn't include the useful categories of the factor; alternatively, we could use a factor variable, in which case adding a new row will take a bit more effort to babysit the factor levels. The option you choose will likely depend on the context of the task at hand. 

## Appending to a data frame 

The key to remember when adding data to a data frame is that columns are vectors and rows are lists. We can also glue two data frames together with ```rbind```:

```{r}
gapminder<-rbind(gapminder, gapminder)
# Having stacked two gapminder datasets on top of each other, 
# we'd now expect twice the number of rows as in the original dataset 
nrow(gapminder)
```

Let's take a look at the dataset:

```{r}
tail(gapminder, n=3)
```

If the row names are unnecessarily complicated and you want to remove them and have R rename them sequentially, you can use the following:

```{r}
rownames(gapminder)<-NULL
head(gapminder)
```

